---
title: "Basic arithmetic in R"
output: 
  learnr::tutorial:
    theme: paper
    highlight: default
runtime: shiny_prerendered
author: Rob Knell
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = TRUE)
```


## Arithmetic operators

R uses the standard arithmetic operators to add, multiply etc. and these are:

| **Operator** | **What it does**      |
|:-------------|:----------------------|
| +            | Addition              |
| -            | Subtraction           |
| *            | Multiply              |
| /            | Divide                |
| ^            | Raise to the power of |

All you have to do to carry out basic arithmetic in R is to type numbers and operators as you might with a calculator. If these are submitted to the console it will print an answer.

```{r}
25 + 3
```

```{r}
8 / 2
```

```{r}
16 ^ 0.5
```
<br><br><br>

### Arithmetic Exercises 

Write the R code required to add twenty seven and six:

```{r 27_plus_6, exercise=TRUE}

```

```{r 27_plus_6-solution}
27 + 6 
```

Write the R code required to subtract eight from nineteen:

```{r 19_minus_8, exercise=TRUE}

```

```{r 19_minus_8-solution}
19 - 8
```

Write the R code required to add calculate 4 raised to the power of three and a half:

```{r 4_to_the_power_of, exercise=TRUE}

```

```{r 4_to_the_power_of-solution}
4 ^ 3.5
```

## Order of calculations

When given some code with more than one operator, R carries out the operations in the order:

1) Parentheses: if something is within brackets (or parentheses) it's done first, then:
1) Exponents (things raised to the power of something), then:
2) Multiply
3) Divide
4) Addition
5) Subtraction in that order.

So `2+3*8` will give the result of `3*8` plus `2`, or `24 + 2`, rather than `3+2` multiplied by `8`. In other words the answer to `2+3*8` is `26`, not `40`. To make R calculate `3+2` and then multiply that by `8` we can use brackets (or parentheses if you're American), so `(3+2)*8` should return a value of 40.

Let's check that.

```{r echo = TRUE}
2 + 3 * 8
```

```{r  echo = TRUE}
(2 + 3) * 8
```

Here's a more complex example. Let's say we want to calculate 

$$  23 + \frac{12 \times 0.66^2}{\left( 1 + 3 \right) ^ {0.5}} $$

If we just put these numbers into R as:

```{r echo = TRUE}
23 + 12 * 0.66^2 / 1 + 3^0.5
```

We get one answer, but it's the wrong one. This is because R is calculating `3^0.5` and then adding 1. What we want to do, as indicated by the brackets in the formula, is add `1+3` and then raise that sum to the power 0.5. We can fix this easily by bracketing this off:

```{r  echo = TRUE}
23 + 12 * 0.66^2 / (1 + 3)^0.5
```

**Top tip** there are lots of helpful mnemonics to make it easier to remember the order of operations, for example **P**lease **E**xcuse **M**y **D**ear **A**unt **S**ally --- **P**arentheses **E**xponents **M**ultiplication **D**ivision **A**ddition **S**ubtraction.

<br><br><br>

### Order of calculations exercises

Use R to calculate

$$ 23 + \frac{\left( 12 \times 0.66 \right)^2}{1 + 3  ^ {0.5}} $$
```{r order1, exercise = TRUE}

```

```{r order1-solution}
23 + (12 * 0.66)^2 / (1 + 3 ^ 0.5)
```

Now use it to calculate

$$ \sqrt{\left( 120 - 16 \right)} + \frac{2 * \left(3*2 \right)}{12} $$


```{r order2, exercise = TRUE}

```

<!-- <div id="order2-hint"> -->
<!-- Don't forget that $5 ^ {0.5} = \sqrt{5}$ -->

<!-- Are all the brackets in the original equation necessary? -->
<!-- </div> -->

```{r order2-hint-1}
# Raising something to the power 0.5 is the same as taking its square root
```

```{r order2-hint-2}

# There are lots of ways of doing this

# Simplest
(120 - 16) ^ 0.5 + 2 * 3 * 2 / 12

# If you want to be sure (when in doubt, brackets are your friend
(120 - 16) ^ 0.5 + (2 * 3 * 2) / 12

# Using sqrt instead of ^0.5
sqrt(120-16) + 2 * 3 * 2/12
```


```{r order2-solution}

# Simplest solution
(120 - 16) ^ 0.5 + 2 * 3 * 2 / 12
```



## Arithmetic Quiz



```{r quiz, echo = FALSE}
quiz(
  question("Which of the following return the answer 25?",
    answer("`19 + 12 * 0.5`", correct = TRUE),
    answer("`(19 + 12) * 0.5`"),
    answer("`19 + (12 * 0.5)`", correct = TRUE),
    answer("`6^2 - 9`")
  ),
  question("Which of the following return a positive answer?",
    answer("`11 * (4 - 11)`"),
    answer("`11 * 4 - 11`", correct = TRUE),
    answer("`(-0.8)^2`", correct = TRUE),
    answer("`-0.8 ^ 2`", message = "This one is a little tricky: R counts the minus sign before 0.8 as an arithmetic operator so it returns the answer -0.64")
  ),
    question("Which of the following is the same as the code `17 + 0.33^2 * 1/5`?",
    answer("$17 + \\sqrt{0.33} \\times \\frac{1}{5}$"),
    answer("`(17 + 0.33 ^2) * 1/5`"),
    answer("$17 + 0.33^{2 + \\frac{1}{5}}$"),
    answer("$17 + 0.33^{2} \\times \\frac{1}{5}$", correct = TRUE),
    answer("$\\left(17 + 0.33\\right) ^ {2} \\times \\frac{1}{5}$"),
    answer("`17 + (0.33 * 0.33) * 1/5`", correct = TRUE)
  )
)
```



